|

### 2023-05-16 paper/00-main.Rmd
more minor edits
more revision

---
title: |
  | \Large Bearing the STRAIN: Analyzing Pass Rush in American Football with Player Tracking Data

# Deforming the line:
# Straining the line:
# Unleashing the Rush
# straining for the takedown
# straining for success
# straining forward
# straining for pressure
# Unleashing the Strain
# Strain on the Gridiron
# from strain to gain
# Bearing the Strain:
# Straining the trenches:

@@

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  # fig.pos = "p",
  cache = TRUE
)
library(here)
library(tidyverse)
```

@@

We thank the organizers of the 2023 NFL Big Data Bowl for hosting the competition and providing access to the data.

>>>>

---
title: |
  | Get Strained: Analyzing Pass Rush in American Football with Player Tracking Data

# Deforming the line:
# Straining the line:
# Unleashing the Rush
# straining for the takedown
# straining for success
# straining forward
# straining for pressure
# Unleashing the Strain
# Strain on the Gridiron
# from strain to gain
# Bearing the Strain:
# Straining the trenches:
# Bearing the STRAIN

@@

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  # fig.pos = "ht",
  cache = TRUE
)
library(here)
library(tidyverse)
```

@@

We thank the organizers of the NFL Big Data Bowl 2023 for hosting the competition and providing access to the data.

### 2023-05-16 paper/01-introduction.Rmd
more revision

In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (as well as the ball).
While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl (BDB) competition [@nfl2023big].

@@

In this paper, we focus on the evaluation of defensive linemen in football.
We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
Our metric gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.
This allows for the assessment of pass rush success even on plays that do not result in an observed outcome like a sack, hit, or hurry.
We view this as a major step forward for accurately evaluating defensive linemen performance.
Further, we note that although our focus in this paper is on defensive linemen, our proposed approach can be extended to the evaluation of offensive linemen in American football.

The remainder of this manuscript is outlined as follows.
We first describe the player tracking data provided by the Big Data Bowl 2023 in Section \ref{sec:data}.
We then introduce the mathematical motivation and definition of our statistic STRAIN and our modeling approach in Section \ref{sec:methods}.
Next, we present applications of STRAIN and study different statistical properties of the metric in Section \ref{sec:results}.
We close with our discussion of future directions related to this work in Section \ref{sec:discussion}.

>>>>

In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (and also the football).
While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl (BDB) competition [@nfl2023big].

@@

In this paper, we focus on the evaluation of defensive linemen in football.
We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
Our statistic gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.
This allows for the assessment of pass rush success even on plays that do not result in an observed outcome like a sack, hit, or hurry.
We view this as a major step forward for accurately evaluating defensive linemen performance.
We also demonstrate that STRAIN is a stable quantity over time and predictive of defensive pressure.
Additionally, we consider a multilevel model to estimate every pass rusher's contribution to the average STRAIN in a play while controlling for other positions like QB and pass blocker as well as various play-level information.
We note that although our focus in this paper is on pass rushers, our proposed approach can be extended to the evaluation of pass blockers in American football.

The remainder of this manuscript is outlined as follows.
We first describe the player tracking data provided by the Big Data Bowl 2023 in Section \ref{sec:data}.
We then introduce the mathematical motivation and definition of our measure STRAIN and our modeling approach in Section \ref{sec:methods}.
Next, we present applications of STRAIN and study different statistical properties of the metric in Section \ref{sec:results}.
We close with our discussion of future directions related to this work in Section \ref{sec:discussion}.

### 2023-05-16 paper/03-methods.Rmd
more revision

Moreover, since we observe STRAIN at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is

\begin{equation}
\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
\end{equation}
This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
We also use average STRAIN to assess different statistical properties of our metric in Section \ref{sec:statprop}.

@@

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language]. Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution <!-- Posterior distribution of what? --> via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].

>>>>

Moreover, since we observe STRAIN at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is
\begin{equation}
\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
\end{equation}
This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
We also use average STRAIN to assess different statistical properties of our metric in Section \ref{sec:statprop}.

@@

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].

### 2023-05-16 paper/04-results.Rmd
more revision

[CHECK/EDIT THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING]

The results of fitting the multilevel model described in Section \ref{sec:multilevel} is displayed in Table \ref{tab:fixedeff}.
First, we investigate the fixed-effects terms of this model.
It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates.
In other words, NFL pass rushers tend to generate more pressure in plays with shorter length and when facing fewer number of blockers, which makes intuitive sense.

@@

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals. 
We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the top pass rusher for their respective position. 
We observe different levels of variability across these top players, with only a subset of players with intervals strictly above zero. This is not necessarily surprising given the limited sample of data. Additionally, we are ignoring correlated performance between pass rushers on the same team.
For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), are teammates on the Cleveland Browns.
McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
We leave the exploration of adjusting for teammates for future work.

>>>>

<!-- [CHECK/EDIT THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING] -->

The results of fitting the multilevel model described in Section \ref{sec:multilevel} is displayed in Table \ref{tab:fixedeff}.
First, we investigate the fixed-effects terms of this model.
It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates.
In other words, NFL pass rushers tend to generate more pressure in plays with shorter length (e.g., blitzes) and when facing fewer number of blockers, which makes intuitive sense.

@@

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the mean varying intercepts for the top ten pass rushers for each position along with the corresponding 95% confidence intervals. 
We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the first-ranked pass rusher for their respective position. 
We observe different levels of variability across these top players, with only a subset of players with intervals strictly above zero. This is not necessarily surprising given the limited sample of data.
Additionally, we are ignoring correlated performance between defensive linemen on the same team.
For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), are teammates on the Cleveland Browns.
McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
We leave the exploration of adjusting for teammates for future work.

### 2023-05-16 paper/05-discussion.Rmd
more revision

We have proposed STRAIN, a simple and interpretable statistic for evaluating pass-rushers, with higher STRAIN corresponds to greater pass-rushing ability. 
Compared to existing metrics, STRAIN measures pass-rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a sack or converting continuous data to a binary win/loss.

As for future directions, while we focus solely on pass-rushers in this paper, STRAIN can be easily extended to the assessment of pass-blockers as a unit. 
This can be accomplished by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.

Summary

Critique/Limitations

Future work rapm

>>>>

In this work, we have proposed STRAIN - a simple and interpretable statistic for evaluating pass rushers, with higher STRAIN values corresponding to greater pass rushing ability. We demonstrated that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events. This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player-tracking data can lead to the development of insightful metrics in American football.

Through multilevel modeling, we observe that the pass rusher explains more variation in play-level STRAIN, followed by the quarterback and then their assigned blocker.
This is an intuitive result, consistent with previous literature on pass rushing and pressure.
However, we ignore the fact that pass rushers can be teammates cooperating in the same play leading to some type of correlation structure in their STRAIN values.
One way to address this complication could be with a adjusted plus-minus (APM) regression approach [@matano2023augmenting], with all players accounted for in modeling a total STRAIN quantity.

In addition, we believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking in American football.
One potential idea is to consider a quantity called stress, which measures force over an area.
By definition, force is the product of mass and acceleration, meaning this quantity would take into account the physical size of a pass rusher in the computation of force.
We could then divide this force over the "area" of the pocket formed by pass blockers to compute stress.

Moreover, although we focus solely on pass rushers in this paper, STRAIN can be easily extended to the assessment of pass blockers as a unit.
This can be accomplished through the aforementioned APM regression or by simply looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
It is also possible to apply STRAIN to assess individual offensive pass blockers, provided that a method of matching blockers to rushers was developed.
Furthermore, compared to existing metrics, STRAIN measures pass rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a binary outcome such as a sack.
There is ample opportunity for working with the complete STRAIN trajectories via functional data analysis to help understand how STRAIN affects offensive production.
